---
title: "Data Importation Practice"
author: "Rylee Elling"
date: "2026-02-04"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_download: true
    code_folding: show

---

# Reading CSV Files

## Locally

Note: Because you chose to use RMarkdown, there could be discrepancy between the path needed for when you run the code line by line and when you run it in RMarkdown.

  - **RMarkdown:** Uses the path based on the RMarkdown file location.
  - **Line-by-line Testing:** Uses the path based on the Project location.
  
Fix: Whenever we open a RMarkdown, we will automatically set the working directory to the folder containing the markdown. 
How to: Go to files on the side, click the settings (ensure you are in the folder you want), select "Set as Working Directory"
Check the Working Directory: write getwd() in console

## From a Local Path

```{r}
usa_df = readr::read_csv("../Data/US000.csv")
```

## From an Online Location

```{r}
unemp_rate <- readr::read_csv("https://fred.stlouisfed.org/graph/fredgraph.csv?bgcolor=%23ebf3fb&chart_type=line&drp=0&fo=open%20sans&graph_bgcolor=%23ffffff&height=450&mode=fred&recession_bars=on&txtcolor=%23444444&ts=12&tts=12&width=1320&nt=0&thu=0&trc=0&show_legend=yes&show_axis_titles=yes&show_tooltip=yes&id=UNRATE&scale=left&cosd=1948-01-01&coed=2025-12-01&line_color=%230073e6&link_values=false&line_style=solid&mark_type=none&mw=3&lw=3&ost=-99999&oet=99999&mma=0&fml=a&fq=Monthly&fam=avg&fgst=lin&fgsnd=2020-02-01&line_index=1&transformation=lin&vintage_date=2026-02-04&revision_date=2026-02-04&nd=1948-01-01")
```
 
Gives us a quick glimpse of the data
```{r}
dplyr::glimpse(unemp_rate)
```


## From GitHUb

You need to get the path for the raw file since that strips away any HTML.

Do this by clicking "Raw" on the github page and then saving the link to the raw page

```{r}
super_ad <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/2e9bd5a67e09b14d01f616b00f7f7e0931515d24/data/2021/2021-03-02/youtube.csv")
```


## JSON Files

JSON stands for JavaScript Object Notation

Most common way of returning results from an API.

Remember to switch working directory to the folder this file is in. 

To read JSON files we will use the **jsonlite** package. 

```{r json_demo}
cincy_jobs = jsonlite::fromJSON(
  txt = "https://www.themuse.com/api/public/jobs?page=1&location=Cincinnati%2C%20OH", flatten = T
)

names(cincy_jobs) # a quick look of what we just read

#we know to look at results by looking at our data and seeing that that is where the good stuff is
cincy_jobs_res = cincy_jobs[['results']]
# cincy_jobs$results does the same thing


#The following is optional

#overwriting the results
cincy_jobs_res = cincy_jobs_res |> tidyr::unnest(
  #specifying cols that are lists and hoping to make them contain single values 
  cols=c('locations', 'categories', 'levels'),
  names_repair = "universal"
  #identified by running dplyr::glimpse(cincy_jobs_res)
) #does seem like we lose data by doing this
# will take extra work or something depending on how you want to analyse the data

#if you want to combine the different rows for that same job that got put in 4 rows in the function above
#you can go by jobid (unique) and then summarise to combine the locations field using commas
# with the constraint of keeping all cols
#dplyr::group_by(id) |> dplyr::summarise(#unfinished
  
#)

```

Note that the choice of converting list cols to separate cols/rows will depend on your data analysis stretch. 



