---
title: "Introduction To R"
author: "Rylee Elling"
date: "2026-02-02"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment operators

In R, there are three assignment operators:

  - '->',  
  - '=',   
  - '<-'.


## R is a Vector Language


```{r}
x_vec = rnorm(n=10, mean=0, sd=1) #generating std normal dist data
x_vec > 0 #finding which elements in x are larger than 0

sum(x_vec > 0) #summing the number of elements (how many are > 0)
```
Looking at the output from the code above, both obtained outputs are vectors.
x_vec > 0 returns a vector of size 10 (T/F for each element)
sum(x_vec > 0) returns a vector of size 1 

## Atomic Vectors: Data Types

There are four data types we will utilize the most in our course:
  - Character (used for text/string)  
  - Double (Real or Decimal)  
  - Integer (No fractions)  
  - Logical (T or F)  
  
We can check the type of an object:
```{r}
typeof(x_vec)
```

```{r}
dept = c('ACC', 'ECO', 'FIN', 'ISA', 'MGMT')
nfaculty_dec = c(18, 19, 14, 25, 22)
nfaculty = c(18L, 19L, 14L, 25L, 22L)
n_mixed = c('string', 13L, 401)
n_mixed_num = c(131L, 402)

typeof(dept)
typeof(nfaculty_dec)
typeof(nfaculty)
typeof(n_mixed)
typeof(n_mixed_num)

print('-----------------')
class(dept)
class(nfaculty_dec)
class(nfaculty)
class(n_mixed)
class(n_mixed_num)
```

## Data Structures: 1D to 2D [In Code]

```{r}
library(tibble)
fsb_tbl <- tibble(
  department = dept,
  count = nfaculty,
  percentage = count / sum(count))
fsb_tbl

```



## Lists

```{r}
lst <- list(
  1:3, # atomic double/numeric vector of length 3
  'a', # atomic character vector of length 1 (aka scalar)
  c(TRUE, FALSE, TRUE), # atomic logical vector of length 3
  c(2.3, 5.9) # atomic double/numeric vector of length = 2
)
lst 

str(lst)
```

A list can contain other lists, i.e. recursive.

```{r}
str(
  list(first_el = lst, second_el= iris)
)
```

```{r}

sublist1 = lst[1]
from_sublist_to_underlying_structure = lst[1] # we get a list
more_subset = lst[[1]] # we get the further underlying structure, in this case an integer
```


## Matrices

We won't use all that often in class.
A matrix is a 2D data structure made of ONE data type.
The length of each of its vectors must be the same, giving it its rectangular structure

```{r}
x_mat = matrix( sample(1:10, size = 4), nrow=2,ncol=2)
str(x_mat)
x_mat # printing nicely
print('-----------------')
x_mat[1,2]
print('-----------------')
x_mat[,2]

```

## Data Frames

A data frame is a named list of vectors with attributes for (column) names, row.names, and its class, “data.frame”.

```{r}
df1 <- data.frame(x = 1:3, y = letters[1:3])
typeof(df1) # showing that its a special case of a list

attributes(df1) # but also is of class data.frame
```

A data frame has an additional constraint: the length of each of its vectors must be the same.

Why are 'Data Frames' generally more useful for data analysis than 'Matrices'?

  - Data frames can store different data types in different columns

## Tibbles

Print easier than data frames. 

Tibbles are designed to be safer and more user-friendly: they print a compact preview, avoid automatic string-to-factor conversion, and are stricter about subsetting (for example, they don’t do partial name matching and are less likely to simplify results unexpectedly).

Tibbles print more rows and columns by default, never convert strings to factors automatically, and use stricter subsetting that is less likely to silently drop types or dimensions. 

# Functions

A function call consists of the function name followed by one or more argument within parentheses.
Function name: mean(), a built-in R funciton to compute mean of a vector

```{r}
temp_high_forecast <- c(87, 84, 85, 89, 9, 84, 81)

```

You can call a function by name or by position

```{r}
mean(x=temp_high_forecast, trim=0.1) # by name
mean(temp_high_forecast, 0.1) # by position
```

Use Functions from Packages

```{r}
#dplyr::cummean(temp_high_forecast)
#dplyr::first(temp_high_forecast)
#dplyr::last(temp_high_forecast)
```




# Subsetting



Atomic vectors are 1D data structures where all elements must have the same type. Since they are 1D, subset using [element_no(s)].

```{r}
x_vec = rnorm(3)
x_vec[c(1,3)]
x_vec[c(1,2,3)]
x_vec[1]
```

A matrix is a 2D data structure made of one/homogeneous data type.

```{r}
x_mat = matrix( sample(1:10, size = 4), nrow = 2, ncol = 2 )
x_mat # printing it nicely
print('-----------------')
x_mat[1, 2] # subsetting
```
```{r}
x_char = matrix( sample(letters, size = 12), nrow = 3, ncol =4 )
x_char
x_char[1:2, 2:3] # subsetting
```

Lists

[ subsets and always returns a list 

[[ extracts a single element from the list

$ is shorthand for extracting a named element (like x[["y"]]).

```{r}
lst <- list( 1:5, "a", c(TRUE, FALSE, TRUE), c(2.3, 5.9) )
lst[4]
lst[[4]]
```

```{r}
my_list <- list(name = "Alice", age = 30, scores = c(85, 90, 78))

my_list[1]
my_list[[1]]
```


Tibbles

Subset tibbles to 1D subset with [[]] or $ 
Subset tibbles by columns with [] or [, col]
Subset tibbles by row with [row,]
Rows and cols with [row,col]

Summary:
Use [[ to extract 1d vectors from 2d tibbles
Use [ to subset tibbles to a new tibble
  - numbers (positive/negative) as indices
  - characters (column names) as indices
  - logicals as indices

```{r}
dept = c('ACC', 'ECO', 'FIN', 'ISA', 'MGMT')
nfaculty = c(18L, 19L, 14L, 25L, 22L)
fsb_tbl <- tibble::tibble(
  department = dept, 
  count = nfaculty, 
  percentage = count / sum(count))
fsb_tbl

print('-----------------')
# subset to 1D
fsb_tbl[["count"]] # column name
fsb_tbl[[2]] # column position
fsb_tbl$count # column name

print('-----------------')
# subset by col
fsb_tbl["count"]
fsb_tbl[2] # for data.frames -> fsb_tbl[, 2]

print('-----------------')
# subset by row
fsb_tbl[c(1, 3), ]
fsb_tbl[-c(2, 4), ]

print('-----------------')
# subset by row and col
fsb_tbl[1:3, 2:3]
## ## fsb_tbl[-4, 2:3] # same as above
## ## fsb_tbl[1:3, c("count". "percentage")] # same result
## ## fsb_tbl[c(rep(TRUE, 3), FALSE), 2:3] # same as above

```

```{r}
fsb_tbl[["count"]] # will produce 1-D vector
fsb_tbl$count # will produce 1D vector
# Resulting in tibbles
fsb_tbl[, 2]
fsb_tbl[1:3, 2:3]
```



## Importing Data

Reading Plain-Text Rectangular

  - delimited text files with read_delim()
    - .csv: comma separated values with read_csv()
    - .tsv: tab separated values read_tsv()
  - .fwf: fixed width files with read_fwf()

```{r}
# when there is an =, that means there is a defaut value
readr::read_csv(
  file,
  col_names = TRUE,
  col_types = NULL,
  locale = default_locale(),
  na = c("", "NA"),
  quoted_na = TRUE,
  quote = "\"",
  comment = "",
  trim_ws = TRUE,
  skip = 0,
  n_max = Inf,
  guess_max = min(1000, n_max),
  progress = show_progress(),
  skip_empty_rows = TRUE
)
```

